---
title: "Functions"
author: "Daijiang Li"
date: "11/15/2022"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

One of the most powerful feature of R is its large packages (~20k). Each package contains some useful functions that allow users to use them to convert some input into an output. Today, we will learn how to write basic R functions.

# Why?

- Functions allow for reproducible code to remove possible tedious and error prone copy-and-paste.
    + If you need to copy and paste some code twice, make it a function! (**Don't repeat yourself** DRY)
- Functions make changing things easy (**Make change in one place**); when we need different parameters, we can just change the arguments of the function.
- Functions can make complex programs tractable by having them performing specific tasks.
- Functions are necessary to develop your own methods/algorithms and share with others.

# The basic structure of a function

The basic structure of a function is:

```r
my_function_name <- function(data, arg1 = value1, arg2 = value2, arg3, arg4, ...){
    <command to do things with the data, with args to control behaviors>
    return(results)
}
```

# Example 1


```{r}
fahrenheit_to_celsius <- function(temp_F) {
  temp_C <- (temp_F - 32) * 5 / 9
  return(temp_C) # return is optional
}
fahrenheit_to_celsius(68)
```

Exercise: write a function to convert celsius to fahrenheit.

```{r}
celsius_to_kelvin <- function(temp_C) {
  temp_K <- temp_C + 273.15
  return(temp_K)
}

# freezing point of water in Kelvin
celsius_to_kelvin(0)
```

Exercise: write a function to convert fahrenheit to kelvin. Hint: use `fahrenheit_to_celsius()` and `celsius_to_kelvin()`.

```r
fahrenheit_to_kelvin <- function(temp_F) {
  temp_C <- ____________
  temp_K <- ____________
  temp_K
}
```

**Tips**: Write modularized functions!

# Example 2

The environment of a function controls how R finds the value associated with a name.

```{r}
f <- function(x) {
  x + y
}
```

R uses rules called lexical scoping to find the value associated with a name. Since y is not defined inside the function, R will look in the environment where the function was defined:

```{r}
y <- 100
f(10)

y <- 1000
f(10)
```


# Example 3

Recall from home work 5 that we wrote some code to extract the binomial names from species names. The `neonDivData` has multiple data frames for different taxonomic groups. How can we make it as a function so that we can apply it to all other data frames?

**Tips**: Write working code first, then convert into a function

```{r}
library(neonDivData)
x = sub(pattern = "^([^ ]+ [^ ]*) .*", replacement = "\\1", x = data_plant$taxon_name)
sample(unique(x), 100)
```


**Naming is hard**: have clear and short function names.

Now, the first step is to think about a name for the function! What should we name it? `get_binomial_name`? Other options? 

**Tip**: Verbs for function names

**Tip**: Be consistent with style: snake_case, camelCase, etc.

```r
get_binomial_name <- function(){

}
```

What about arguments? What kind of arguments should the function have??

**Tip**: Nouns for arguments.


```r
get_binomial_name <- function(v){

}
```

Exercise: finish the above function.




```r
get_binomial_name <- function(v){
  v2 = sub(pattern = "^([^ ]+ [^ ]*) .*", replacement = "\\1", x = data_plant$taxon_name)
  return(v2)
}
```

What's wrong with the above function??


How can we improve the function?

```{r}
get_binomial_name <- function(v, patterns = "^([^ ]+ [^ ]*) .*"){
  v2 = sub(pattern = patterns, replacement = "\\1", x = v)
  return(v2)
}

get_binomial_name <- function(v, 
                              patterns = "^([^ ]+ [^ ]*) .*",
                              repl = "\\1"
                              ){
  v2 = sub(pattern = patterns, replacement = repl, x = v)
  return(v2)
} # benefit?? be more flexible?
```

**Tips:** Data arguments first; detail arguments later with default values.

```{r}
head(get_binomial_name(data_bird$value), 10)
head(data_bird$value, 10)
class(data_bird$value)
```

**Tips:** Defensive coding and stop early; or return early.

**If a function is going to fail, fail early!**

```r
get_binomial_name <- function(v, 
                              patterns = "^([^ ]+ [^ ]*) .*",
                              repl = "\\1"
                              ){
  if(!is.character(v)){
    stop("The input vector is not character.")
  }
  v2 = sub(pattern = patterns, replacement = repl, x = v)
  return(v2)
} 
get_binomial_name(data_bird$value)
# Error in get_binomial_name(data_bird$value) : 
#   The input vector is not character.
```

**Sometimes, return early too!**

```r
f <- function(args){
    if(x){
      some complex calculation
      out = results
    } else {
      some simple calculation
      out = results
    }
    return(out)
}

# return early to save time
f <- function(args){
    if(!x){
    some simple calculation
    out = results
    return(out)
    }
    
    some complex calculation
    results
}
```


**Tips:** Use comments to explain why.


```{r}
get_binomial_name <- function(v, 
                              patterns = "^([^ ]+ [^ ]*) .*",
                              repl = "\\1"
                              ){
  if(!is.character(v)){ # in case non-character input
    stop("The input vector is not character.")
  }
  v2 = sub(pattern = patterns, replacement = repl, x = v)
  return(v2)
} 
```

**Tips:** `...`

```{r}
get_binomial_name <- function(v, 
                              patterns = "^([^ ]+ [^ ]*) .*",
                              repl = "\\1",
                              ...
                              ){
  if(!is.character(v)){ # in case non-character input
    stop("The input vector is not character.")
  }
  v2 = sub(pattern = patterns, replacement = repl, x = v, ...)
  return(v2)
} 
head(get_binomial_name(data_bird$taxon_name, perl = TRUE), 50)
```


**Tips:** Document arguments! (use `{roxygen2}`)

```{r}
#' To extract binomial scientific names
#' 
#' Some description about the function.
#'
#' @param v Input vector, must be a character vector.
#' @param patterns The Regex pattern to match, default will extract the first two words.
#' @param repl The values to replace the pattern matched that was specified by `patterns`.
#' @param ... Additional arguments to be passed to `[sub]` function.
#' @return A character vector, with the binomial scientific names
#' @examples get_binomial_name(c("Carex aquatilis Wahlenb.", "Boerhavia coulteri (Hook. f.) S. Watson"))
#'
get_binomial_name <- function(v, 
                              patterns = "^([^ ]+ [^ ]*) .*",
                              repl = "\\1",
                              ...
){
  if(!is.character(v)){ # in case non-character input
    stop("The input vector is not character.")
  }
  v2 = sub(pattern = patterns, replacement = repl, x = v, ...)
  return(v2)
} 
sample(unique(get_binomial_name(data_fish$taxon_name)), 30)
sample(unique(get_binomial_name(data_small_mammal$taxon_name)), 30)
```

**Note**: Avoid overriding existing functions and variables.










