---
title: "if...else..."
author: "Daijiang Li"
date: "9/30/2021"
output: 
  html_document: 
    toc: yes
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Control flow

Sometimes, we only want to execute some statements if certain conditions are met; sometimes, we want to execute some statement repetitively for each elements of a dataset or until certain conditions/values reached. This is where control-flow constructs come in. In this week's lectures, we will largely focus on this topic.

For the syntax examples throughout this lecture, keep the following in mind:

- *statement* is a single R statement or a compound statement (a group of R statements enclosed in curly braces `{ }` and separated by semicolons).
- `cond` is an expression that resolves to `TRUE` or `FALSE`.
- `expr` is a statement that evaluates to a number or character string.
- `seq` is a sequence of numbers or character strings.

## Conditional execution

In conditional execution, a statement or statements are only executed if a specified condition is met. These constructs include `if-else`, `ifelse`, and `switch`.

Before we talk about these functions, it is necessary to talk about R's logical operators.

|   Operator |             Description |
| ---------: | ----------------------: |
|        `<` |               less than |
|       `<=` |   less than or equal to |
|        `>` |             larger than |
|       `>=` | larger than or equal to |
|       `==` |        exactly equal to |
|        `!` |             opposite of |
|       `!=` |            not equal to |
|        `|` |                      or |
|        `&` |                     and |
|     `%in%` |                      in |
| `isTRUE()` |             is it true? |

[**Surprise surprise!**](https://stackoverflow.com/questions/9508518/why-are-these-numbers-not-equal)

```{r eval=FALSE}
i = 0.1
i = i + 0.05
i == 0.15 # ? what do you say?
```

