---
title: "if...else..."
author: "Daijiang Li"
date: "9/19/2023"
output: 
  html_document: 
    toc: yes
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Control flow

Sometimes, we only want to execute some statements if certain conditions are met; sometimes, we want to execute some statement repetitively for each elements of a dataset or until certain conditions/values reached. This is where control-flow constructs come in. In this week's lectures, we will largely focus on this topic.

For the syntax examples throughout this lecture, keep the following in mind:

- *statement* is a single R statement or a compound statement (a group of R statements enclosed in curly braces `{ }` and separated by semicolons).
- `cond` is an expression that resolves to `TRUE` or `FALSE`.
- `expr` is a statement that evaluates to a number or character string.
- `seq` is a sequence of numbers or character strings.

## Conditional execution

In conditional execution, a statement or statements are only executed if a specified condition is met. These constructs include `if-else`, `ifelse`, and `switch`.

Before we talk about these functions, it is necessary to talk about R's logical operators.

|   Operator |             Description |
| ---------: | ----------------------: |
|        `<` |               less than |
|       `<=` |   less than or equal to |
|        `>` |             larger than |
|       `>=` | larger than or equal to |
|       `==` |        exactly equal to |
|        `!` |             opposite of |
|       `!=` |            not equal to |
|        `|` |                      or |
|        `&` |                     and |
|     `%in%` |                      in |
| `isTRUE()` |             is it true? |

[**Surprise surprise!**](https://stackoverflow.com/questions/9508518/why-are-these-numbers-not-equal)

```{r eval=FALSE}
i = 0.1
i = i + 0.05
i == 0.15 # ? what do you say?
```

# `if ... else ...`

How does an if else statement work in R?

- Two steps: evaluation and execution (which are not necessarily repeated)
- Two outcomes in evaluation
    - Condition satisfied (i.e., outcome is TRUE) -> execute commands of "if"
    - Condition not satisfied (i.e., outcome is FALSE) -> execute commands of "else" (if any)
- if else statements can be nested


```{r eval=FALSE}
# General form
if(cond){
  statement for TRUE
}

# else is optional
if(cond){
  statement for TRUE
} else {
  statement for FALSE
} 
# bracket can be ignored too; 
# but I find it is easier to read with it

if(cond){
  statement for TRUE
} else {
  another if ... else ...
} 
```

```{r}
a <- 4
if (a %% 2 == 0) {
  b1 <- a * -1
}
a
b1
```

```{r}
a <- 5
if (a %% 2 == 0) {
  b <- a * -1
} else {
  b <- a ^ 2
}
b
```

```{r}
# nested if else
a <- 0.33
if (a %% 2 == 0) {
  b <- a * -1
} else {
  if (a %% 2 == 1) {
    b <- a ^ 2
  } else {
    b <- a * 10
  }
}
b
```

Why use if else statement?

- Useful for situations with multiple scenarios
- Define your own categories
- Case-by-case manipulation of data

```{r}
# Define your own categories
small <- 1:5
medium <- 6:10
large <- 11:15

a <- 2
if (a %in% small) {
  b <- 'Small number'
} else if (a %in% medium) {
  b <- 'Medium Number'
} else if (a %in% large) {
  b <- 'Large number'
} else {
  b <- 'Number out of range'
}
b

a <- 12
if (a %in% small) {
  b <- 'Small number'
} else {
  if (a %in% medium) {
    b <- 'Medium Number'
  } else {
    if (a %in% large) {
      b <- 'Large number'
    } else {
      b <- 'Number out of range'
    }
  }
} 
b
```

#### Defensive coding

When write your own function or doing analysis, we should be very careful about potential errors / failures. It is recommended to warn / message / fail early!

```{r}
if(TRUE) warning("This is a warning")
if(TRUE) message("This is a message")
# if(TRUE) stop("Stoped because of ...")
stopifnot()
```


So if else is useful to work with one value/situation/condition/evaluation at each time.  What if we have multiple values/cases need to evaluate? 

- `ifelse()`
- combine with loops such as `for()` (next lecture)

# `ifelse()`

`ifelse()` is a compact and vectorized version of the if-else construct. The syntax is:

```{r eval=FALSE}
ifelse(cond, statement for TRUE, statement for FALSE)
```

### Problem to solve

Suppose that we have a vector `vec_a` with 5 elements:

```{r}
vec_a <- c(3, 2, -5, 7, 0)
```

Now we want to take square root for *positive and zero* values while leave negative values alone. How can we do this?

We may do the following things step by step for **each element**:

1. Is it non-negative?
2. If yes, take the square root (recall the functions from last class).
3. If no, return the original value.

#### Solution 1: combine if else with for()

We will talk more about loops in the next lecture.


```{r}
vec_a2 <- vector(mode = "numeric", length = length(vec_a)) # to hold results
for (i in 1:length(vec_a)){
  if (vec_a[i] >= 0){
    vec_a2[i] = sqrt(vec_a[i])
  } else {
    vec_a2[i] = vec_a[i]
  }
} 
vec_a2
```

|           Pros |                                                                                                  Cons |
| -------------: | ----------------------------------------------------------------------------------------------------: |
|      Intuitive |   Multiple steps of coding: reate vector to hold results, then the 3-steps procedure for each element |
| Easy to follow |                                                        Relatively slow (imagine millions of elements) |


#### Solution 2: use `ifelse()`

```{r}
ifelse(vec_a >= 0, sqrt(vec_a), vec_a)
vec_a2
```


#### Vectorization

A "whole project" thinking instead of "element-wise".

- In `R`, this means a function takes the whole vector as input (instead of one element each time the `if()` function takes within the `for()` loop above)
- In `R`, vectorized functions are much faster
- But `for` loops are still very important (see next class)
- [More information about vectorization](http://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html)

**Why warning message??**

```r
Warning message:
In sqrt(vec_a) : NaNs produced
```

```{r}
(vec_test <- vec_a >= 0)
(vec_yes <- sqrt(vec_a)) # warning here
(vec_no <- vec_a)
ifelse(vec_test, vec_yes, vec_no) # no warning
```

- if the *i*th element of `vec_test` is `TRUE`
- then take the *i*th element of `vec_yes`
- otherwise, take the *i*th element of `vec_no`

Now, suppose instead of return the original value if it is negative, we want to return `NA`.

```{r}
ifelse(vec_a >= 0, sqrt(vec_a), NA)
sqrt(ifelse(vec_a >= 0, vec_a, NA)) # No warning any more
```

#### `ifelse()` works for whole matrix too

```{r}
mat_a <- matrix(data = c(1:6, -2, -5, 0), nrow = 3, ncol = 3)
mat_a
ifelse(mat_a > 0, sqrt(mat_a), mat_a)
```

#### More examples

Suppose we have exam scores for students, and we want to assign pass (>= 60) or fail (<60) to each of them.

```{r}
set.seed(123)
scores <- data.frame(names = letters,
                     score = runif(n = 26, min = 30, max = 100))
head(scores, n = 3)
hist(scores$score)
```

How do we add another column that will indicate pass and fail?

```{r}
str(scores)
scores$score
(pass_fail <- ifelse(scores$score >= 60, "pass", "fail")) # recycle
scores$status <- pass_fail
head(scores, n = 6)
table(scores$status)
table(scores$score >= 60)
```

What about if we want also assign "A" (>= 85), "B" (>= 75), "C" (>= 60) to those who passed the exam?

```{r}
# nested ifelse
vec_levels <- 
  ifelse(scores$score < 60, "fail", 
       ifelse(scores$score < 75, "C",
              ifelse(scores$score < 85, "B", "A")))
vec_score <- scores$score # to make it easier to check
names(vec_score) <- vec_levels # in pratice, you should make it as a new column
vec_score

dplyr::case_when(
  scores$score < 60 ~ "fail",
  scores$score < 75 ~ "C",
  scores$score < 85 ~ "B",
  TRUE ~ "A"
)
```

If needed, you can also assign "D" (e.g. >= 50, < 60), "E" (< 50) to those who failed by replacing `"fail"` with another `ifelse()`.

```{r fig.width = 6, fig.asp=1, out.height=400, fig.align='center'}
# "A" (>= 85), "B" (>= 75), "C" (>= 60)
boxplot(vec_score ~ vec_levels)
abline(h = c(60, 75, 85), lty = 2)
```

**Be careful when use `ifelse()` with Dates and factors**

```{r}
(vec_x <- factor(c("b", "a", "d", "e")))
(vec_y <- ifelse(vec_x == "a", vec_x, NA))
as.numeric(vec_x)
vec_x2 <- as.character(vec_x) # convert first
ifelse(vec_x2 == "a", vec_x2, NA)
```

See `?ifelse` for example of Dates

- The results of `ifelse()` have the same length as the input `test`, but attributes may not been preserved
- Values are selected from `vec_yes` and `vec_no`
- If `vec_yes` and `vec_no` are too short, they are recycled
- `NA` in the input give `NA` in the output

## `if else ` vs `ifelse()`

`if(test) yes else no` is much more efficient and often much preferable to `ifelse(test, yes, no)` whenever `test` is a simple true/false result, i.e., when `length(test) == 1`.


# `switch()`

`switch` chooses statements based on the value of an expression. The syntax is `switch(expr, ...)` where `...` represents statements tied to the possible outcome values of `expr`. It’s easiest to understand how switch works by looking at the example in the following listing.

```{r}
feelings <- c("sad", "afraid")
for (i in feelings){
  print(
    switch(i,
           happy  = "I am glad you are happy",
           afraid = "There is nothing to fear",
           sad    = "Cheer up",
           angry  = "Calm down now"
    ) 
  )
}
```

`switch` can be very useful when write our own functions to deal with different scenarios. 

```{r}
mydate <- function(type = "long") {
  switch(type,
         long =  format(Sys.time(), "%A %B %d %Y"),
         short = format(Sys.time(), "%m-%d-%y"),
         cat(type, "is not a recognized type\n")
  ) 
}
mydate("long")
mydate("short")
mydate("medium")
```

