---
title: "Loops"
author: "Daijiang Li"
date: "10/05/2021"
output: 
  html_document: 
    toc: yes
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this lecture, we will talk about loops. Loops are very powerful in terms to do some tasks repeatedly (like thousands of times). Computers are very good with doing repeated tasks, which we as human are pretty bad. Even though we present examples in R here, the logic applies to pretty much any programming language.

Loops repeat two steps: evaluation and execution. For evaluation, we tell the computer to test for a specific condition, and there are two outcomes of this evaluation: condition satisfied and not satisfied. Then for each scenario, we can let the computer know what to do respectively.

Three types of loops in R:

- `repeat` 
- `while` 
- `for` 

# `repeat` 

`repeat` is the simplest loop to repeat the same expression

```{r eval=FALSE}
repeat {expression}
```

To stop repeating the expression, you can use the keyword `break`. To skip to the
next iteration in a loop, you can use the command `next`.

```{r}
i <- 5
repeat {
  if (i > 25) 
    break 
  else 
    print(i); i <- i + 5
}
```

If you do not include a `break` command, the R code will be an infinite loop.

# `while`

Another useful construction is `while` loops, which repeat an expression while a condition is true:

```{r eval=FALSE}
while (condition) expression
```

```{r}
i <- 5
while (i <= 25) { 
  print(i) 
  i <- i + 5
}
```

You can also use `break` and `next` inside while loops. The `break` statement is used to stop iterating through a loop. The `next` statement skips to the next loop iteration
without evaluating the remaining expressions in the loop body.


# `for` 

```{r eval=FALSE}
# syntax
for (variable in sequence){
  statements for each variable
}
```

**Results are not printed inside a loop unless you explicitly call the `print` function.**

```{r}
for (i in 1:5) {
  paste(i, "^2 = ", i ^ 2, sep = "")
}
```

Add `print`.

```{r}
for (i in 1:5) {
  print(paste(i, "^2 = ", i ^ 2, sep = ""))
}
```

```{r}
for (i in 1:5) {
  # Concatenate and Print
  cat(i, "^2 = ", i ^ 2, "\n", sep = "")
}
```

```{r}
a <- c(2, 6)
for (i in 1:2) {
  print(paste('i', '=', i, sep=' ')) 
  print(paste(c('a', '=', a), collapse=' ')) 
  cat('execution\n') 
  a[i] <- a[i] + 10 
  print(paste(c('a', '=', a), collapse=' ')) 
  print(paste('i', '=', i, sep=' ')) 
  cat('next\n\n') 
}
a
a <- c(2, 6) 
a + 10
```

**The variable `var` that is set in a for loop is changed in the calling environment.**

```{r}
b <- character(2) # to hold results
i <- 1
for (i in 1:2) {
  b[i] <- toupper(a[i])
}
i
b
toupper(a)
```

```{r}
years <- 2016:2025
for (i in seq_along(years)) {
  print(paste('Year', years[i], sep=' '))
}
paste('Year', 2016:2025, sep=' ')
```

# Applying a Function to Each Element of an Object

Except loops such as `for()`, The base R packages include a set of different functions to apply a function to each element (or a subset of elements) of an object.

## To apply a function to part of an array or a matrix

To apply a function to parts of an array (or matrix), use the `apply` function:

```{r eval=FALSE}
apply(array, margin, FUN, ...)
# ... are arguments to be passed to FUN
```

```{r}
x = matrix(1:20, nrow = 5)
x
# for each row, find the max value
for(i in 1:nrow(x)){
  print(max(x[i, ]))
}
# apply
apply(X = x, MARGIN = 1, FUN = max, nam.rm = TRUE)

# for each column, find the max value
for(i in 1:ncol(x)){
  print(max(x[, i]))
}
# apply
apply(X = x, MARGIN = 2, FUN = max, nam.rm = TRUE)
```

A more complicated example:

```{r}
x = array(1:27, dim = c(3, 3, 3))
x
apply(X = x, MARGIN = 1, FUN = paste, collapse = ", ")
apply(X = x, MARGIN = 2, FUN = paste, collapse = ", ")
apply(X = x, MARGIN = 3, FUN = paste, collapse = ", ")
apply(X = x, MARGIN = c(1, 2), FUN = paste, collapse = ", ")
```
```{r include=FALSE, eval=FALSE}
# for loop
x2 = matrix(0, nrow = 3, ncol = 3)
for(i in 1:nrow(x2)){
  for(j in 1:ncol(x2)){
    vv = vector(length = 3)
    for(k in 1:3){
      vv[k] = x[i, j, k]
    }
    x2[i, j] = paste(vv, collapse = ", ")
  }
}
```

## Apply a function to a list or vector

To apply a function to each element in a vector or a list and **return a list**, you can use the function `lapply`. The function `lapply` requires two arguments: an object `X` and a function `FUN`. (You may specify additional arguments that will be passed to `FUN`.)

```{r}
x <- 1:5
lapply(X = x, FUN = function(i) 2^i)

x <- data.frame(x = 1:5, y = 6:10)
lapply(X = x, FUN = function(i) 2^i)
lapply(X = x, FUN = mean)
```

Sometimes, you might prefer to get a vector, matrix, or array instead of a list. To do this, use the `sapply` function. This function works exactly the same way as `lapply`,
except that it returns a vector or matrix (when appropriate).

```{r}
sapply(X = x, FUN = function(i) 2^i)
vapply(X = x, FUN = function(i) 2^i, FUN.VALUE = rep(1, 5))
```


Another related function is `mapply`, the “multivariate” version of `sapply`.

```{r}
mapply(paste, 
       c(1, 2, 3, 4, 5),
       c("a","b","c","d","e"),
       c("A","B","C","D","E"),
       MoreArgs = list(sep = "-")
)
```

# Tidyverse




