---
title: "The file system and basic unix shell"
subtitle: "Introduction to Data Science (BIOL7800)\nhttps://introdatasci.dlilab.com/"
author: "Daijiang Li"
institute: "LSU"
date: "2021/09/21"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts", "../style.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align='center', out.width = '95%')
```

class: middle

### A file system is a logical collection of files on a partition or disk

#### Unix filesystem (common folders):

|     Directory    |                                Description                               |
|:----------------:|:------------------------------------------------------------------------:|
|        `/`       | Root directory                                                           |
|      `/bin`      | Executable files for all users                                           |
|      `/dev`      | Device drivers                                                           |
|      `/home`     | Home directory for users                                                 |
|      `/tmp`      | Temporary files used between system boots                                |
|      `/usr`      | Used for miscellaneous purposes, e.g. admin commands, shared files, etc. |
| `/Users` (macOS) | Files of different users                                                 |


???

a disk can have multiple partitions

---

# File tree

```bash
# Debian / mint / Ubuntu
sudo apt-get install tree
# macOS & linux (with homebrew installed)
brew install tree
```

.pull-left[

```bash
tree -L 3 .
```

```
.
├── libs
│   ├── header-attrs
│   │   └── header-attrs.js
│   └── remark-css
│       ├── default-fonts.css
│       └── default.css
├── presentation.Rmd
└── presentation.html

3 directories, 5 files
```
]

.pull-right[

```{r}
fs::dir_tree(path = ".") #R
```

]

---

# Naming files and directories

.font150[
- All names are case sensitive
- You can use pretty much any character except `/`
    + **Better to avoid space in names!** (looking at you, `Box Sync`!)
    + Also try to avoid special characters such as `>`, `|`, `:`, `&`, `!`
    + `\` backslash-escaping the character (i.e. quote it). `ls ~/Box\ Sync`
    + files and directories start with `.` will be hidden
- Names are generally limited to 255 characters
- Use date format ISO: YYYYMMDD or YYYY-MM-DD
]

---

# Naming files and directories

.font150[
- Name should be consistent (machine friendly) and descriptive (human readable)
    + `data.csv` (not meaningful) 
    + `Li_plant_diversity_WI_2020-06-08.csv` (better)
- Name files so they can ordered 
    + `00_pkg_functions.R`
    + `01_data.R`
- Have a file extension that matches the file format
]

## .blue[Take time to practice and adopt these styles and strategies can save you a lot of time in the future]

---

# Absolute vs. relative paths

`pwd` to check where you are right now. Let's assume we are at `/Users/dli/IntroDataScience/lectures/06_bash`

.pull-left[
### Absolute path

```bash
# list files within the current folder
ls /Users/dli/IntroDataScience/lectures/06_bash

# move one level up?
cd /Users/dli/IntroDataScience/lectures 

# move two levels up?
cd /Users/dli/IntroDataScience 

# move to the home folder?
cd /Users/dli # /home/user
```
]


.pull-right[
### .blue[Relative path]

```bash
# list files within the current folder
ls .

# move one level up?
cd ..

# move two levels up?
cd ../..

# move to the home folder?
cd ~
```
]


---


# Some basic commands

.font150[
- `man` to read the manual! Press `q` to quit, `space` to read next page (`--help` of no `man`)
- `Tab` for auto-completence
- Creating, removing, changing directories
    + `mkdir`
    + `rmdir`
    + `cd`
- Moving, coping, removing files
    + `touch` to create empty files
    + `mv`
    + `cp`
    + `rm`: notable options: `-r`, `-f` (dangerous `rm -rf`)
]

---

class: center, middle, inverse

.font200[.red[Warning:] no undo button on the command line, so be extremely careful with deleting files/directories! I recommend to use `mv` to move files to the Trash Bin instead. (`~/.Trash` on Mac, `~/.local/share/Trash` on Ubuntu)]

---

# Some basic commands

.font150[
- List files with `ls`: `-a`, `-l`
- Wildcard to work with multiple files and folders
    + `ls /bin/*sh`
    + `ls /bin/*[ab]*`
    + `mv /path/to/*.jpeg /new/path`
- Check history of used command `history`
    + `!200` to rerun the 200th command from history
- `open` (macOS) or `start`
]

---

# The Unix Philosophy

### Each program should **do one thing well**

### Programs can be linked with the pipe operator `|` to do more complex tasks

```bash
ls -l | head -n 3

head -100 file.csv | tail -20 | > file_81_100.csv
```

---

## Moving cursor inside bash

.font200[
+ `^` means `Ctrl` key, e.g. `^A` means `Ctrl+A`.
+ `^A` move to the beginning of a line in the shell.
+ `^E` move to the end of a line.
+ `^C` cancel what you are doing. If it does not work, try `^\`.
+ `^D` end of a line. This normally will kill the job too.
+ `^L` clean the screen of your shell.
+ `^h` delete back one character (backspace).
+ `^w` delete back one word.
+ `^u` delete back to the start of line.
]

---

class: middle, inverse

## More common commands see [course webpage](https://introdatasci.dlilab.com/2021-08-18-shell-commands/)
